Should geophysicists use gravity anomaly or gravity disturbance?

### 2018-05-11 manuscript/letter.tex

\author[Hallam et al.]
{Kristoffer A. T. Hallam$^{1}$\thanks{Emails: kristoffer.hallam@gmail.com,
vanderlei@on.br}, Vanderlei C. Oliveira Jr$^1$,
Val\'{e}ria C. F. Barbosa$^1$ \linebreak and Leonardo Uieda$^2$ \\
$^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
$^2$ Department of Geology and Geophysics, University of Hawaii, Manoa, USA
}

@@

The resultant of gravitational force and centrifugal force acting 
on a body at rest on the Earth's surface is called 
\textit{gravity vector} and its intensity is called 
\textit{gravity} \citep{heiskanen-moritz1967,
hofmann-wellenhof-moritz2005}.
In the case of gravimetry on moving platforms (e.g., airplanes,
helicopters, marine vessels), there are additional
non-gravitational accelerations due to the vehicle motion, 
such as Coriolis acceleration and high-frequency vibrations 
\citep{glennie-etal2000,nabighian-etal2005-grav,baumann-etal2012}.
Geophysicists use gravity for estimating the Earth's 
internal density distribution whereas geodesists use
gravity to estimate the Geoid \citep{li2001}.
Hence, geophysicists are usually interested 
in the gravitational component of the observed gravity, 
which is produced by the Earth's internal 
density distribution.

For isolating this 
gravitational component consists in removing the non-gravitational 
effects due to the vehicle motion and also the time variations 
such as Earth tides, instrumental drift and barometric 
pressure changes, for example.
If these effects are properly removed, the resultant 
gravity data can be considered as the sum of a 
centrifugal component due to the Earth's rotation and
a gravitational component produced by the whole Earth's
internal density distribution.
The isolation of this particular gravitational component 
and its subsequent use for estimating density 
distributions related to geological structures in subsurface 
are the main goals in applied geophysics \citep{blakely1996}.
The computation of the gravitational effect produced by
the geological structures is commonly known in geophysics as 
\textit{gravity modelling}.
This term is different from \textit{gravity field modelling}, which
is commonly used in geodesy to denote the characterization of the 
gravity field in a local, regional or global scales.

Based on well-established concepts of the literature,
we present a discussion aiming at critically examining
the following question: in geophysical applications,
should we use the gravity disturbance or gravity anomaly?
It seems that this theoretical issue has been 
debated within the scientific community from a 
more geodetic than geophysical point of view
\citep{lafehr1991,chapin1996,li2001,fairhead2003,
hackney-featherstone2003,hinze2005}.
Our reasoning suggests that the gravity 
disturbance is more appropriated than the gravity anomaly 
for geophysical purposes.

@@

The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$, at the same 
point $P$, defines a quantity called \textit{gravity disturbance 
vector}, which is given by:
\begin{equation}
\delta \mitbf{g}_{P} = 
\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}
Because the centrifugal part of the normal gravity vector 
is equal to the centrifugal component of the gravity vector, 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
a purely gravitational effect caused by contrasts between the actual internal 
density distribution of the Earth and the internal density 
distribution of the normal Earth.
ergo, the gravity disturbance vector $\delta \mitbf{g}_{P}$ 
(eq. \ref{eq:gravity-disturbance-vector}) is harmonic.
It is important to stress that both density distributions are unknown.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, 
\citeyear{forsberg1984}) or \textit{gravity sources} (e.g., 
\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for 
using the last term.
Fig. \ref{fig:surfaces} illustrates the gravity vector 
$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$ 
located on the Earth's surface.

The difference between the magnitudes of gravity vector
$g_{P} = \| \mitbf{g}_{P} \|$ and the normal gravity vector
$\gamma_{P} = \| \mitbf{\gamma}_{P} \|$, at the 
same point $P$, is called \textit{gravity disturbance} 
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005} 
and can be represented as follows:
\begin{equation}
\delta g_{P} = g_{P} - \gamma_{P} \: .
\label{eq:gravity-disturbance}
\end{equation}
Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
to the magnitude of the gravity disturbance vector 
$\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.

@@

\end{document}\grid

>>>>

\author[Oliveira Jr et al.]
{Vanderlei C. Oliveira Jr$^1$\thanks{Email: vanderlei@on.br},
Kristoffer A. T. Hallam$^{1}$, Val\'{e}ria C. F. Barbosa$^1$
and Leonardo Uieda$^2$ \\
$^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
$^2$ Department of Geology and Geophysics, SOEST, University of Hawai'i at M\={a}noa, Honolulu, USA
}

@@

The \textit{gravity vector} is the sum of the gravitational and centrifugal
accelerations felt by a body at rest on the Earth's surface.
Its intensity is what is geoscientists call \textit{gravity}
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}.
When measured on moving platforms (e.g., airplanes,
helicopters, marine vessels), there are additional
non-gravitational accelerations due to the motion of the vehicle,
such as the Coriolis acceleration and high-frequency vibrations
\citep{glennie-etal2000,nabighian-etal2005-grav,baumann-etal2012}.
Geodesists use gravity measurements to estimate the Geoid \citep{li2001}
whereas geophysicists use gravity to estimate the Earth's
internal density distribution.
Hence, geophysicists are usually only interested
in the gravitational component of the observed gravity
because that is what that reflects the Earth's internal density distribution.

The first step toward isolating the gravitational component
is to remove the effects of vehicle motion, Earth tides, instrumental drift,
and barometric pressure changes, among others.
If these effects are properly removed,
the resultant observations are considered to be solely
the sum of the centrifugal acceleration due to the Earth's rotation and
the gravitational acceleration produced by
the internal density distribution of the whole Earth.
The isolation of this particular gravitational component,
and its subsequent use for estimating density
distributions related to geological structures in subsurface,
are the main goals of applied gravimetry \citep{blakely1996}.
The computation of the gravitational effect produced by
the geological structures is known in geophysics as
\textit{gravity modelling}.
Not to be confused with \textit{gravity field modelling}, which
is commonly used in geodesy to denote the mathematical characterization of the
gravity field in local, regional, or global scales (e.g., spherical harmonic
analysis).

We present a discussion with the aim of shedding some light on the question:
\textit{should we use the gravity disturbance or gravity anomaly in geophysical
applications}?
The concepts discussed here are well-established in the literature.
However, the debate around this theoretical issue has been
carried about from a more geodetic than a geophysical point of view
\citep{lafehr1991,chapin1996,li2001,fairhead2003,
hackney-featherstone2003,hinze2005}.
Our reasoning suggests that the gravity disturbance is more appropriated for
geophysical purposes than the gravity anomaly.

@@

The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$, at the same 
point $P$, defines a quantity called \textit{gravity disturbance 
vector}, which is given by:
\begin{equation}
\delta \mitbf{g}_{P} = 
\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}
Because the centrifugal part of the normal gravity vector 
is equal to the centrifugal component of the gravity vector, 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
a purely gravitational effect caused by contrasts between the actual internal 
density distribution of the Earth and the internal density 
distribution of the normal Earth.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, 
\citeyear{forsberg1984}) or \textit{gravity sources} (e.g., 
\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for 
using the last term.
Fig. \ref{fig:surfaces} illustrates the gravity vector 
$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$ 
located on the Earth's surface.

The difference between the magnitudes of gravity vector
$g_{P} = \| \mitbf{g}_{P} \|$ and the normal gravity vector
$\gamma_{P} = \| \mitbf{\gamma}_{P} \|$, at the 
same point $P$, is called \textit{gravity disturbance} 
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005} 
and can be represented as follows:

\begin{equation}
\delta g_{P} = g_{P} - \gamma_{P} \: .
\label{eq:gravity-disturbance}
\end{equation}

Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
to the magnitude of the gravity disturbance vector 
$\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.

@@

\end{document}
